---
title: "Peak cherry bloom prediction"
author: "Nick Pullen"
date: "`r lubridate::now()`"
output: 
  html_document:
    code_folding: hide
    keep_md: false
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
knit: (function(input, ...) {
    rmarkdown::render(
      input, 
      output_format = 'all',
      output_file = paste0(
        xfun::sans_ext(input), '_', format(Sys.time(), "%Y-%m-%d-%H%M")),
      envir = new.env(),
      output_dir = here::here("output")
    )
  })
---
<style>
pre {
 overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
div.main-container {
  width: 100%;
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>
```{r setup, results='show', message=TRUE, warning=TRUE}
knitr::opts_chunk$set(message = FALSE, echo = TRUE)
options(width = 300)
library(tidyverse)
library(lubridate)
library(flextable)
library(Hmisc)
library(rnoaa)

# Functions
myflextable = function(data, ...) {
  set_flextable_defaults(na_str = "NA", theme_fun = theme_booktabs, font.size = 12, padding.bottom = 1, padding.top = 1)
  x = flextable(data, ...)
  x = colformat_int(x, big.mark = "")
  x = colformat_double(x, big.mark = "", digits = 2, na_str = "NA")
  return(x)
}
```

## Load data
```{r}
dc = read_csv("data/washingtondc.csv")
liestal = read_csv("data/liestal.csv")
swiss = read_csv("data/meteoswiss.csv") %>% 
  mutate(location = str_remove(location, "Switzerland/"))
kyoto = read_csv("data/kyoto.csv")
japan = read_csv("data/japan.csv")
korea = read_csv("data/south_korea.csv")
```

## Check quantiles especially
```{r}
Hmisc::describe(dc)
head(ch)
tail(kyoto)
quantile(kyoto$bloom_doy, probs = c(0, 0.1, 0.25, 0.4, 0.5, 0.6, 0.75, 0.9, 1))
quantile(ch$bloom_doy, probs = c(0, 0.1, 0.25, 0.4, 0.5, 0.6, 0.75, 0.9, 1))
quantile(dc$bloom_doy, probs = c(0, 0.1, 0.25, 0.4, 0.5, 0.6, 0.75, 0.9, 1))
```

```{r}
liestal %>% 
  ggplot(aes(x = year, y = bloom_doy)) +
  geom_point(shape = 21, fill = 'plum1') +
  geom_line(colour = 'plum1') +
  labs(x = "Year", y = "Peak bloom (days since Jan 1st)")
```

```{r}
swiss %>% 
  ggplot(aes(x = year, y = bloom_doy, colour = location)) +
  geom_point() +
  facet_wrap(~location) +
  labs(x = "Year", y = "Peak bloom days since Jan 1st") + 
  theme(legend.position = "none")
```

## Climate data
```{r}
#' @param stationid the `rnoaa` station id (see [ghcnd_stations()]) # go to https://www1.ncdc.noaa.gov/pub/data/ghcn/daily/ghcnd-stations.txt if this Timeouts: Error in curl::curl_fetch_disk(x$url$url, x$disk, handle = x$url$handle) : Timeout was reached: [ftp.ncdc.noaa.gov] Connection timed out after 10001 milliseconds
#' @return a data frame with columns
#'   - `year` ... the year of the observations
#'   - `season` ... the season (Winter, Spring, Summer, Fall)
#'   - `tmax_avg` ... average maximum temperate in tenth degree Celsius
get_temperature = function(stationid) {
  ghcnd_search(stationid = stationid, var = c("tmax", "tmin"), 
               date_min = "1950-01-01", date_max = "2022-01-31", refresh = TRUE)[[1]] %>% # [[1]] is tmax, so do 2 for tmin or rbind separately
    mutate(year = as.integer(format(date, "%Y")),
           month = as.integer(strftime(date, '%m')) %% 12, # make December "0"
           season = cut(month, breaks = c(0, 2, 5, 8, 11),
                        include.lowest = TRUE,
                        labels = c("Winter", "Spring", "Summer", "Fall")),
           year = if_else(month == 0, year + 1L, year)) %>%
    group_by(year, season) %>%
    summarise(tmax_avg = mean(tmax, na.rm = TRUE))#,
              #tmin_avg = mean(tmin, na.rm = TRUE))
}

historic_temperatures =
  tibble(location = "washingtondc", get_temperature("USC00186350")) %>%
  bind_rows(tibble(location = "liestal", get_temperature("GME00127786"))) %>% # SZ000001940 = BASEL BINNINGEN
  bind_rows(tibble(location = "kyoto", get_temperature("JA000047759"))) %>%
  bind_rows(tibble(location = "vancouver", get_temperature("CA001108395")))

historic_temperatures %>%
  filter(location == "kyoto" & season == "Spring") %>%
  ggplot() + 
  aes(year, tmax_avg) + 
  geom_line() +
  xlim(1951, 2031) +
  labs(x = "Year", y = "Average maximum temperature (1/10 °C)")
```

## Integrated Surface Data - Lite (Hourly temp data)
https://www1.ncdc.noaa.gov/pub/data/noaa/isd-lite/
From the Technical Document
 > The ISD-Lite product is designed to be an easier to work with subset of the larger Integrated Surface Data hourly dataset.  ISD-Lite  > contains eight common hourly time-series climatological variables represented in a fixed-width format.  The elements extracted are:
 > 
 > 1.	Air temperature (degrees Celsius * 10)
 > 2.	Dew point temperature (degrees Celsius * 10)
 > 3.	Sea level pressure (hectopascals)
 > 4.	Wind direction (angular degrees)
 > 5.	Wind speed (meters per second * 10)
 > 6.	Total cloud cover (coded, see format documentation)
 > 7.	One-hour accumulated liquid precipitation (millimeters)
 > 8.	Six-hour accumulated liquid precipitation (millimeters)
 > 
 > The ISD-Lite data are represented with a modified time stamp which corresponds to the nearest hour of actual observation.  Sub-hourly  > observations were removed.  Duplicate observations were resolved according to a ranking system.  
 
Also note
> the absence of flag data may make it difficult to determine good data from suspicious data

Data format is:
 - Year Month Day Hour AirTemp*10  ......
 - MISSING VALUE: -9999
 
```{r}
stations <- isd_stations()
wloc = stations %>% filter(str_detect(station_name, "BASEL|RONALD R|KYOTO|VANCOUVER INTL"),
                     str_detect(end, "2022"))
  #  usaf   wban  station_name                     ctry  state icao     lat     lon elev_m    begin      end
  # <chr>  <chr> <chr>                            <chr> <chr> <chr>  <dbl>   <dbl>  <dbl>    <dbl>    <dbl>
  # 066010 99999 BASEL-BINNINGEN                  SZ    ""    ""      47.6    7.58  317   19901101 20220210
  # 477590 99999 KYOTO                            JA    ""    ""      35.0  136.     46.5 19451031 20220210
  # 718920 99999 VANCOUVER INTL                   CA    ""    "CYVR"  49.2 -123.      4.3 19550702 20220210
  # 724050 13743 RONALD REAGAN WASHINGTON NATL AP US    "VA"  "KDCA"  38.8  -77.0     4.1 19360901 20220211
```

```{r}

my_isd_remote = function (usaf, wban, year) 
{
    file.path(rnoaa:::isdbase(), "isd-lite", year, sprintf("%s-%s-%s%s", usaf, 
        wban, year, ".gz"))
}
my_isd_local = function (usaf, wban, year, path, ext) 
{
    file.path(path, sprintf("%s-%s-%s%s", usaf, wban, year, 
        ext))
}
my_isd_GET = function (usaf, wban, year, overwrite, force, ...) 
{
    isd_cache$mkdir()
    rds <- my_isd_local(usaf, wban, year, isd_cache$cache_path_get(), 
        ".rds")
    if (!rnoaa:::is_isd(rds) || force) {
        fp <- my_isd_local(usaf, wban, year, isd_cache$cache_path_get(), 
            ".gz")
        cli <- crul::HttpClient$new(my_isd_remote(usaf, wban, year), 
            opts = list(...))
        #browser()
        tryget <- tryCatch(suppressWarnings(cli$get(disk = fp)), 
            error = function(e) e)
        if (inherits(tryget, "error") || !tryget$success()) {
            unlink(fp)
            stop("download failed for\n   ", my_isd_remote(usaf, 
                wban, year), call. = FALSE)
        }
    }
    return(rds)
}
my_isd = function(usaf, wban, year, overwrite = TRUE, cleanup = TRUE,
                   additional = TRUE, parallel = FALSE, cores = getOption(
                     "cl.cores",
                     2
                   ), progress = FALSE, force = FALSE, ...) {
  rds_path <- my_isd_GET(usaf, wban, year, overwrite, force, ...)
  gz_path <- sub("rds", "gz", rds_path)
  # browser()
  if (!file.exists(gz_path)) {
    stop("file not found", call. = FALSE)
  }
  message(sprintf("<path>%s", gz_path), "\n")

  lns <- readLines(gz_path, encoding = "latin1")
  linesproc <- vector(mode = "list", length = length(lns))
  for (i in seq_along(lns)) {
    linesproc[[i]] <- lns[i]
  }
  df <- as_tibble(unlist(linesproc))
  saveRDS(df, rds_path)
  if (cleanup) {
    unlink(gz_path) # delete
  }
  attr(df, "source") <- rds_path
  df
}
```

## Get Basel weather data
```{r}
my_isd(usaf = "066010", wban = "99999", year=1990)

basel_years = year(ymd(wloc %>% filter(ctry=="SZ") %>% pull(begin))):year(ymd(wloc %>% filter(ctry=="SZ") %>% pull(end)))
basel_list = vector(mode = "list", length = length(basel_years))

#basel_years = c(1990,1991)
for (i in seq_along(basel_years)) {
  tryCatch({
    print(basel_years[i])
    basel_list[[i]] = my_isd(usaf = "066010", wban = "99999", year=basel_years[i])
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

basel_tbl = map_dfr(basel_list, bind_rows) %>%
  separate(value, into = c("yr", "mon", "day", "hr", "temp", rep(NA, 7)), sep = "\\s+", convert = TRUE)
basel_tbl = basel_tbl %>% mutate(datetime = make_datetime(yr, mon, day, hr)) %>% filter(between(temp, -900, 900)) %>% mutate(temp = temp/10)
ggplot(basel_tbl, aes(x=datetime, y=temp)) + geom_line() + facet_wrap(~yr, ncol=4, scales = "free_x")
```

## Get Kyoto weather data
```{r}
kyoto_years = year(ymd(wloc %>% filter(ctry=="JA") %>% pull(begin))):year(ymd(wloc %>% filter(ctry=="JA") %>% pull(end)))
kyoto_list = vector(mode = "list", length = length(kyoto_years))

for (i in seq_along(kyoto_years)) {
  tryCatch({
    print(kyoto_years[i])
    kyoto_list[[i]] = my_isd(usaf = "477590", wban = "99999", year=kyoto_years[i])
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

kyoto_tbl = map_dfr(kyoto_list, bind_rows) %>%
  separate(value, into = c("yr", "mon", "day", "hr", "temp", rep(NA, 7)), sep = "\\s+", convert = TRUE)
kyoto_tbl = kyoto_tbl %>% mutate(datetime = make_datetime(yr, mon, day, hr)) %>% filter(between(temp, -900, 900)) %>% mutate(temp = temp/10)
ggplot(kyoto_tbl, aes(x=datetime, y=temp)) + geom_line() + facet_wrap(~yr, ncol=6, scales = "free_x")
```

## Get Washington DC weather data
```{r}
washingtonDC_years = year(ymd(wloc %>% filter(ctry=="US") %>% pull(begin))):year(ymd(wloc %>% filter(ctry=="US") %>% pull(end)))
washingtonDC_list = vector(mode = "list", length = length(washingtonDC_years))

for (i in seq_along(washingtonDC_years)) {
  tryCatch({
    print(washingtonDC_years[i])
    washingtonDC_list[[i]] = my_isd(usaf = "724050", wban = "13743", year=washingtonDC_years[i])
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

washingtonDC_tbl = map_dfr(washingtonDC_list, bind_rows) %>%
  separate(value, into = c("yr", "mon", "day", "hr", "temp", rep(NA, 7)), sep = "\\s+", convert = TRUE)
washingtonDC_tbl = washingtonDC_tbl %>% mutate(datetime = make_datetime(yr, mon, day, hr)) %>% filter(between(temp, -900, 900)) %>% mutate(temp = temp/10)
ggplot(washingtonDC_tbl, aes(x=datetime, y=temp)) + geom_line() + facet_wrap(~yr, ncol=6, scales = "free_x")
```

## Get Washington DC weather data
```{r}
vancouver_years = year(ymd(wloc %>% filter(ctry=="CA") %>% pull(begin))):year(ymd(wloc %>% filter(ctry=="CA") %>% pull(end)))
vancouver_list = vector(mode = "list", length = length(vancouver_years))

for (i in seq_along(vancouver_years)) {
  tryCatch({
    print(vancouver_years[i])
    vancouver_list[[i]] = my_isd(usaf = "718920", wban = "99999", year=vancouver_years[i])
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

vancouver_tbl = map_dfr(vancouver_list, bind_rows) %>%
  separate(value, into = c("yr", "mon", "day", "hr", "temp", rep(NA, 7)), sep = "\\s+", convert = TRUE)
vancouver_tbl = vancouver_tbl %>% mutate(datetime = make_datetime(yr, mon, day, hr)) %>% filter(between(temp, -900, 900)) %>% mutate(temp = temp/10)
ggplot(vancouver_tbl, aes(x=datetime, y=temp)) + geom_line() + facet_wrap(~yr, ncol=6, scales = "free_x")
```

```{r}
saveRDS(basel_tbl, "data/small_basel_hrly_temp.rds")
saveRDS(kyoto_tbl, "data/small_kyoto_hrly_temp.rds")
saveRDS(washingtonDC_tbl, "data/small_washingtonDC_hrly_temp.rds")
saveRDS(vancouver_tbl, "data/small_vancouver_hrly_temp.rds")
```

